A Comparison of R Tools for Nonlinear Least Squares Modeling

### 2023-05-17 JNABsub20230515/ComparingNLStools.tex
add submission may 23

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={A comparison of R tools for nonlinear least squares modeling},
  pdfauthor={John C. Nash ; Arkajyoti Bhattacharjee },
  colorlinks=true,
  linkcolor={red},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={blue},
  pdfcreator={LaTeX via pandoc}}

\title{A comparison of R tools for nonlinear least squares modeling}
\author{John C. Nash
\thanks{ retired professor, Telfer School of Management, University of Ottawa} \and Arkajyoti
Bhattacharjee
\thanks{Department of Mathematics and Statistics, Indian Institute of Technology, Kanpur}}
\date{2023-3-2}

\begin{document}
\maketitle

\hypertarget{abstract}{%
\subsection{Abstract}\label{abstract}}

Our Google Summer of Code project ``Improvements to \texttt{nls()}''
considered the features and limitations of the R function \texttt{nls()}
with the aim of improving and rationalizing R tools for nonlinear
regression. The rich features of \texttt{nls()} are weakened by several
deficiencies and inconsistencies such as a lack of stabilization of the
Gauss-Newton solver. Further considerations are the usability and
maintainability of the code base that provides the functionality
\texttt{nls()} claims to offer. Various packages, including our
\texttt{nlsr}, provide alternative capabilities. We consider the
differences in goals, approaches, and features of different tools for
nonlinear least squares modeling in R. Discussion of these matters is
relevant to improving R generally as well as its nonlinear estimation
tools.

@@

\texttt{nls()} is the tool for estimating nonlinear statistical models
in base R, the primary software distribution from the Comprehensive R
Archive Network (\emph{\url{https://cran.r-project.org}}). The function
dates to the 1980s and the work related to Bates and Watts
(\protect\hyperlink{ref-bateswatts}{1988}) in S (see
\emph{\url{https://en.wikipedia.org/wiki/S_\%28programming_language\%29}}).

@@

The tools we will consider are the base-R \texttt{nls()} function and
some packages that are available on the CRAN repository. Of these, we
will pay particular attention to \textbf{nlsr} (John C Nash and Duncan
Murdoch (\protect\hyperlink{ref-nlsr2023manual}{2023})),
\textbf{minpack.lm} (Elzhov et al.
(\protect\hyperlink{ref-minpacklm12}{2012})), and \textbf{gslnls} (Chau
(\protect\hyperlink{ref-gslnls23}{2023})) which are general nonlinear
least-squares solvers.

While we will provide capsule comments for some other CRAN packages, we
will not pursue those in the Bioconductor (Gentleman et al.
(\protect\hyperlink{ref-Bioconductor}{2004})) collection, nor those on
repositories such as GitHub (\url{https://github.com}) and Gitlab
(\url{https://about.gitlab.com}).

@@

\texttt{onls} (Andrej-Nikolai Spiess
(\protect\hyperlink{ref-R-onls-manual}{2022})) -- The usual optimization
in estimating nonlinear models is the vertical difference between the
dependent variable and the functional model. \texttt{onls} minimizes the
sum of \textbf{orthogonal} residuals. The objective is therefore
different and involves nontrivial extra calculation. A vignette with the
package and the blog article
\url{https://www.r-bloggers.com/2015/01/introducing-orthogonal-nonlinear-least-squares-regression-in-r/}
give some description with illustrative graphs. \texttt{onls} appears to
be limited to problems with one independent and one dependent variable.
The Wikipedia article
\url{https://en.wikipedia.org/wiki/Total_least_squares} presents an
overview of some ideas, with references to the literature. The approach
needs a wider discussion and tutorial examples to allow its merits to be
judged than can be included here.

\texttt{crsnls} (Tvrdík (\protect\hyperlink{ref-R-crsnls-manual}{2016}))
-- This package allows nonlinear estimation by controlled random search
via two methods. There is unfortunately no vignette. A modest trial we
carried out showed \texttt{nlsr::nlxb()} gave the same results in a
small fraction of the time required by either of the methods in
\texttt{crsnls}. The method discussed in Josef Tvrdík and Ivan Křivý and
Ladislav Mišík (\protect\hyperlink{ref-tvrdik2007}{2007}) claims better
reliability in finding solutions than a Levenberg-Marquardt code
(actually from Matlab), but the tests were conducted on the extreme NIST
examples mentioned next.

\texttt{NISTnls} (National Institutes for Standards and R port by
Douglas Bates (\protect\hyperlink{ref-NISTnls-manual}{2012})) -- This
package provides R code and data for a set of (numerically
ill-conditioned) nonlinear least squares problems from the U.S. National
Institute for Standards and Technology. These may not represent
real-world situations.

\texttt{nlshelper} (Duursma
(\protect\hyperlink{ref-nlshelper-manual}{2017})) -- This package, which
unfortunately lacks a vignette, provides a few utilities for
summarizing, testing, and plotting non-linear regression models
estimated with nls(), nlsList() or nlme() that are linked or grouped in
some way.

\texttt{nlsic} (Sokol (\protect\hyperlink{ref-nlsic-manual}{2022})) --
This solves nonlinear least squares problems with optional equality
and/or inequality constraints. It is clearly \textbf{not} about
modeling, and the input and output are quite different from class
\texttt{nls} methods. However, there do not appear to be other R
packages with these capabilities.

\texttt{nlsMicrobio} (Baty and Delignette-Muller
(\protect\hyperlink{ref-nlsMicrobio-manual}{2014})) -- Data sets and
nonlinear regression models dedicated to predictive microbiology,
including a vignette, by authors of the \texttt{nlstools} package.

\texttt{nlstools} (Baty and Delignette-Muller
(\protect\hyperlink{ref-p-nlstools}{2013})) -- This package provides
several tools for aiding the estimation of nonlinear models,
particularly using \texttt{nls()}. The vignette is actually a journal
article, and the authors have considerable experience in the subject.

\texttt{nlsmsn} (Prates, Lachos, and Garay
(\protect\hyperlink{ref-m-nlsmsn}{2021})) -- Fit univariate non-linear
scale mixture of skew-normal(NL-SMSN) regression, with details in Garay,
Lachos, and Abanto-Valle (2011). The problem here is to minimize an
objective that is modified from the traditional sum of squared
residuals.

\texttt{nls.multstart} (Padfield and Matheson
(\protect\hyperlink{ref-nlsmultstart}{2020})) -- Non-linear least
squares regression using AIC scores with the Levenberg-Marquardt
algorithm using multiple starting values for increasing the chance that
the minimum found is the global minimum.

\texttt{nls2} (Grothendieck (\protect\hyperlink{ref-nls2-22}{2022})) --
Nonlinear least squares by brute force has similar motivations to
\texttt{nls.multstart}, but uses \texttt{nls()} within multiple trials.
The author has extensive expertise in R.

\texttt{nlstac} (Rodriguez-Arias et al.
(\protect\hyperlink{ref-m-nlstac}{2020})) -- A set of functions
implementing the algorithm described in Fernandez Torvisco et al.~(2018)
for fitting separable nonlinear regression curves. The special class of
problem for which this package is intended is an important and difficult
one. No vignette is provided, unfortunately.

@@

\texttt{nlraa} (Miguez (\protect\hyperlink{ref-MiguezNLRAA2021}{2021}))
-- a set of nonlinear \emph{selfStart} models, primarily from
agriculture. Most include analytic Jacobian code.

\texttt{optimx} (John C. Nash and Varadhan
(\protect\hyperlink{ref-p-optimx}{2011})) -- This provides optimizers
that can be applied to minimize a nonlinear function which could be a
nonlinear sum of squares. NOT generally recommended if nonlinear least
squares programs can be easily used, but provides a check and
alternative solvers.

@@

The Hobbs weed infestation problem (John C. Nash
(\protect\hyperlink{ref-cnm79}{1979, 120})) is a growth curve modeling
task that seems straightforward but is quite nasty. Its very succinct
statement provides the ``short reproducible example'' much requested on
R mailing lists. This is a real problem from a field researcher. The
data and graph follow.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{weed }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\FloatTok{5.308}\NormalTok{, }\FloatTok{7.24}\NormalTok{, }\FloatTok{9.638}\NormalTok{, }\FloatTok{12.866}\NormalTok{, }\FloatTok{17.069}\NormalTok{, }\FloatTok{23.192}\NormalTok{, }\FloatTok{31.443}\NormalTok{,}
          \FloatTok{38.558}\NormalTok{, }\FloatTok{50.156}\NormalTok{, }\FloatTok{62.948}\NormalTok{, }\FloatTok{75.995}\NormalTok{, }\FloatTok{91.972}\NormalTok{)}
\NormalTok{tt }\OtherTok{\textless{}{-}} \DecValTok{1}\SpecialCharTok{:}\DecValTok{12}
\NormalTok{weeddf }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(tt, weed)}
\FunctionTok{plot}\NormalTok{(weeddf, }\AttributeTok{main=}\StringTok{"Hobbs weed infestation data"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

In reality, there are many problems were \(D\) should not be ignored,
but the work to compute it precisely is considerable. Many work-arounds
have been proposed, of which the Levenberg-Marquardt stabilization
(Levenberg (\protect\hyperlink{ref-Levenberg1944}{1944}), Marquardt
(\protect\hyperlink{ref-Marquardt1963}{1963})) is the most commonly
used. For convenience, we will use ``Marquardt'', as we believe he first
incorporated the ideas into a practical computer program.

The usual suggestion is that \(D\) be replaced by a multiple of the unit
matrix or else a multiple of the diagonal part of \(J' J\). In low
precision, some elements of \(J' J\) could underflow to zero (John C.
Nash (\protect\hyperlink{ref-jn77ima}{1977})), and a linear combination
of both choices is an effective compromise. Various choices for \(D\),
as well as a possible line search along the direction \(\delta\) rather
than a unit step (Hartley (\protect\hyperlink{ref-Hartley1961}{1961})),
give rise to several variant algorithms. ``Marquardt's method'' is a
family of methods. Fortunately, most choices work well.

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# model formulas}
\NormalTok{frmu }\OtherTok{\textless{}{-}}\NormalTok{ weed }\SpecialCharTok{\textasciitilde{}}\NormalTok{ b1}\SpecialCharTok{/}\NormalTok{(}\DecValTok{1}\SpecialCharTok{+}\NormalTok{b2}\SpecialCharTok{*}\FunctionTok{exp}\NormalTok{(}\SpecialCharTok{{-}}\NormalTok{b3}\SpecialCharTok{*}\NormalTok{tt))}
\NormalTok{frms }\OtherTok{\textless{}{-}}\NormalTok{ weed }\SpecialCharTok{\textasciitilde{}} \DecValTok{100}\SpecialCharTok{*}\NormalTok{c1}\SpecialCharTok{/}\NormalTok{(}\DecValTok{1}\SpecialCharTok{+}\DecValTok{10}\SpecialCharTok{*}\NormalTok{c2}\SpecialCharTok{*}\FunctionTok{exp}\NormalTok{(}\SpecialCharTok{{-}}\FloatTok{0.1}\SpecialCharTok{*}\NormalTok{c3}\SpecialCharTok{*}\NormalTok{tt))}
\NormalTok{frmt }\OtherTok{\textless{}{-}}\NormalTok{ weed }\SpecialCharTok{\textasciitilde{}}\NormalTok{ Asym }\SpecialCharTok{/}\NormalTok{(}\DecValTok{1} \SpecialCharTok{+} \FunctionTok{exp}\NormalTok{((xmid}\SpecialCharTok{{-}}\NormalTok{tt)}\SpecialCharTok{/}\NormalTok{scal))}
\CommentTok{\#}
\CommentTok{\# Starting parameter sets}
\NormalTok{stu1}\OtherTok{\textless{}{-}}\FunctionTok{c}\NormalTok{(}\AttributeTok{b1=}\DecValTok{1}\NormalTok{, }\AttributeTok{b2=}\DecValTok{1}\NormalTok{, }\AttributeTok{b3=}\DecValTok{1}\NormalTok{)}
\NormalTok{sts1}\OtherTok{\textless{}{-}}\FunctionTok{c}\NormalTok{(}\AttributeTok{c1=}\DecValTok{1}\NormalTok{, }\AttributeTok{c2=}\DecValTok{1}\NormalTok{, }\AttributeTok{c3=}\DecValTok{1}\NormalTok{)}
\NormalTok{stt1}\OtherTok{\textless{}{-}}\FunctionTok{c}\NormalTok{(}\AttributeTok{Asym=}\DecValTok{1}\NormalTok{, }\AttributeTok{xmid=}\DecValTok{1}\NormalTok{, }\AttributeTok{scal=}\DecValTok{1}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Logistic3U}
\NormalTok{hobbs.res  }\OtherTok{\textless{}{-}}  \ControlFlowTok{function}\NormalTok{(x)\{ }\CommentTok{\# scaled Hobbs weeds problem {-}{-} residual}
  \CommentTok{\# This variant uses looping}
  \ControlFlowTok{if}\NormalTok{(}\FunctionTok{length}\NormalTok{(x) }\SpecialCharTok{!=} \DecValTok{3}\NormalTok{) }\FunctionTok{stop}\NormalTok{(}\StringTok{"hobbs.res {-}{-} parameter vector n!=3"}\NormalTok{)}
\NormalTok{  y  }\OtherTok{\textless{}{-}}  \FunctionTok{c}\NormalTok{(}\FloatTok{5.308}\NormalTok{, }\FloatTok{7.24}\NormalTok{, }\FloatTok{9.638}\NormalTok{, }\FloatTok{12.866}\NormalTok{, }\FloatTok{17.069}\NormalTok{, }\FloatTok{23.192}\NormalTok{, }\FloatTok{31.443}\NormalTok{, }
           \FloatTok{38.558}\NormalTok{, }\FloatTok{50.156}\NormalTok{, }\FloatTok{62.948}\NormalTok{, }\FloatTok{75.995}\NormalTok{, }\FloatTok{91.972}\NormalTok{)}
\NormalTok{  tt  }\OtherTok{\textless{}{-}}  \DecValTok{1}\SpecialCharTok{:}\DecValTok{12}
\NormalTok{  res  }\OtherTok{\textless{}{-}}\NormalTok{  x[}\DecValTok{1}\NormalTok{]}\SpecialCharTok{/}\NormalTok{(}\DecValTok{1}\SpecialCharTok{+}\NormalTok{x[}\DecValTok{2}\NormalTok{]}\SpecialCharTok{*}\FunctionTok{exp}\NormalTok{(}\SpecialCharTok{{-}}\NormalTok{x[}\DecValTok{3}\NormalTok{]}\SpecialCharTok{*}\NormalTok{tt)) }\SpecialCharTok{{-}}\NormalTok{ y}
\NormalTok{\}}

\NormalTok{hobbs.jac  }\OtherTok{\textless{}{-}}  \ControlFlowTok{function}\NormalTok{(x) \{ }\CommentTok{\# scaled Hobbs weeds problem {-}{-} Jacobian}
\NormalTok{  jj  }\OtherTok{\textless{}{-}}  \FunctionTok{matrix}\NormalTok{(}\FloatTok{0.0}\NormalTok{, }\DecValTok{12}\NormalTok{, }\DecValTok{3}\NormalTok{)}
\NormalTok{  tt  }\OtherTok{\textless{}{-}}  \DecValTok{1}\SpecialCharTok{:}\DecValTok{12}
\NormalTok{  yy  }\OtherTok{\textless{}{-}}  \FunctionTok{exp}\NormalTok{(}\SpecialCharTok{{-}}\NormalTok{x[}\DecValTok{3}\NormalTok{]}\SpecialCharTok{*}\NormalTok{tt)}
\NormalTok{  zz  }\OtherTok{\textless{}{-}}  \FloatTok{1.0}\SpecialCharTok{/}\NormalTok{(}\DecValTok{1}\SpecialCharTok{+}\NormalTok{x[}\DecValTok{2}\NormalTok{]}\SpecialCharTok{*}\NormalTok{yy)}
\NormalTok{  jj[tt,}\DecValTok{1}\NormalTok{]   }\OtherTok{\textless{}{-}}\NormalTok{   zz}
\NormalTok{  jj[tt,}\DecValTok{2}\NormalTok{]   }\OtherTok{\textless{}{-}}   \SpecialCharTok{{-}}\NormalTok{x[}\DecValTok{1}\NormalTok{]}\SpecialCharTok{*}\NormalTok{zz}\SpecialCharTok{*}\NormalTok{zz}\SpecialCharTok{*}\NormalTok{yy}
\NormalTok{  jj[tt,}\DecValTok{3}\NormalTok{]   }\OtherTok{\textless{}{-}}\NormalTok{   x[}\DecValTok{1}\NormalTok{]}\SpecialCharTok{*}\NormalTok{zz}\SpecialCharTok{*}\NormalTok{zz}\SpecialCharTok{*}\NormalTok{yy}\SpecialCharTok{*}\NormalTok{x[}\DecValTok{2}\NormalTok{]}\SpecialCharTok{*}\NormalTok{tt}
  \FunctionTok{attr}\NormalTok{(jj, }\StringTok{"gradient"}\NormalTok{) }\OtherTok{\textless{}{-}}\NormalTok{ jj}
\NormalTok{  jj}
\NormalTok{\}}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{unls1}\OtherTok{\textless{}{-}}\FunctionTok{try}\NormalTok{(}\FunctionTok{nls}\NormalTok{(}\AttributeTok{formula=}\NormalTok{frmu, }\AttributeTok{start=}\NormalTok{stu1, }\AttributeTok{data=}\NormalTok{weeddf))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Error in nls(formula = frmu, start = stu1, data = weeddf) : 
##   singular gradient
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{snls1}\OtherTok{\textless{}{-}}\FunctionTok{try}\NormalTok{(}\FunctionTok{nls}\NormalTok{(}\AttributeTok{formula=}\NormalTok{frms, }\AttributeTok{start=}\NormalTok{sts1, }\AttributeTok{data=}\NormalTok{weeddf))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Error in nls(formula = frms, start = sts1, data = weeddf) : 
##   singular gradient
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{tnls1}\OtherTok{\textless{}{-}}\FunctionTok{try}\NormalTok{(}\FunctionTok{nls}\NormalTok{(}\AttributeTok{formula=}\NormalTok{frmt, }\AttributeTok{start=}\NormalTok{stt1, }\AttributeTok{data=}\NormalTok{weeddf))}
\end{Highlighting}
\end{Shaded}

>>>>

% !TeX root = RJwrapper.tex
\title{A Comparison of R Tools for Nonlinear Least Squares Modeling}
\author{by John C. Nash and Arkajyoti Bhattacharjee}

\maketitle

\abstract{%
Our Google Summer of Code project ``Improvements to \texttt{nls()}'' considered the features and limitations of the R function \texttt{nls()} with the aim of improving and rationalizing R tools for nonlinear regression. The rich features of \texttt{nls()} are weakened by several deficiencies and inconsistencies such as a lack of stabilization of the Gauss-Newton solver. Further considerations are the usability and maintainability of the code base that provides the functionality \texttt{nls()} claims to offer. Various packages, including our \texttt{nlsr}, provide alternative capabilities. We consider the differences in goals, approaches, and features of different tools for nonlinear least squares modeling in R. Discussion of these matters is relevant to improving R generally as well as its nonlinear estimation tools.
}

@@

\texttt{nls()} is the tool for estimating nonlinear statistical models in base R, the
primary software distribution
from the Comprehensive R Archive Network (\emph{\url{https://cran.r-project.org}}).
The function dates to the 1980s and the work related to Bates and Watts (1988) in S (see
\emph{\url{https://en.wikipedia.org/wiki/S_\%28programming_language\%29}}).

@@

The tools we will consider are the base-R \texttt{nls()} function
and some packages that are available on the CRAN repository.
Of these, we will pay particular attention to \textbf{nlsr} (John C Nash and Duncan Murdoch (2023)),
\textbf{minpack.lm} (Elzhov et al. (2012)), and \textbf{gslnls} (Chau (2023)) which are
general nonlinear least-squares solvers.

While we will provide capsule comments for some other CRAN packages,
we will not pursue those in the Bioconductor (Gentleman et al. (2004))
collection, nor those on repositories such as
GitHub (\url{https://github.com}) and Gitlab (\url{https://about.gitlab.com}).

@@

\texttt{onls} (Andrej-Nikolai Spiess (2022)) -- The usual optimization in estimating nonlinear models
is the vertical difference between the dependent variable and the functional
model. \texttt{onls} minimizes the sum of \textbf{orthogonal} residuals. The objective is
therefore different and involves nontrivial extra calculation. A vignette with
the package and the blog article
\url{https://www.r-bloggers.com/2015/01/introducing-orthogonal-nonlinear-least-squares-regression-in-r/}
give some description with illustrative graphs. \texttt{onls} appears to be limited to
problems with one independent and one dependent variable. The Wikipedia article
\url{https://en.wikipedia.org/wiki/Total_least_squares} presents an overview of some
ideas, with references to the literature. The approach needs a wider
discussion and tutorial examples to allow its merits to be judged than can be
included here.

\texttt{crsnls} (Tvrdík (2016)) -- This package allows nonlinear estimation by
controlled random search via two methods. There is unfortunately no vignette.
A modest trial we carried out showed \texttt{nlsr::nlxb()} gave the same results in
a small fraction of the time required by either of the methods in \texttt{crsnls}. The
method discussed in Josef Tvrdík and Ivan Křivý and Ladislav Mišík (2007) claims better reliability in finding solutions
than a Levenberg-Marquardt code (actually from Matlab), but the tests were conducted
on the extreme NIST examples mentioned next.

\texttt{NISTnls} (National Institutes for Standards and R port by Douglas Bates (2012)) -- This package provides R code and data for a set of
(numerically ill-conditioned) nonlinear least squares problems from the U.S. National
Institute for Standards and Technology. These may not represent real-world situations.

\texttt{nlshelper} (Duursma (2017)) -- This package, which unfortunately lacks a
vignette, provides a few utilities for summarizing, testing, and plotting
non-linear regression models estimated with nls(), nlsList() or nlme() that
are linked or grouped in some way.

\texttt{nlsic} (Sokol (2022)) -- This solves nonlinear least squares problems with optional
equality and/or inequality constraints. It is clearly \textbf{not} about modeling, and
the input and output are quite different from class \texttt{nls} methods. However, there do
not appear to be other R packages with these capabilities.

\texttt{nlsMicrobio} (Baty and Delignette-Muller (2014)) -- Data sets and nonlinear regression models
dedicated to predictive microbiology, including a vignette, by authors of the
\texttt{nlstools} package.

\texttt{nlstools} (Baty and Delignette-Muller (2013)) -- This package provides several tools for aiding the estimation
of nonlinear models, particularly using \texttt{nls()}. The vignette is actually a journal
article, and the authors have considerable experience in the subject.

\texttt{nlsmsn} (Prates, Lachos, and Garay (2021)) -- Fit univariate non-linear scale mixture of skew-normal(NL-SMSN)
regression, with details in Garay, Lachos, and Abanto-Valle (2011). The problem
here is to minimize an objective that is modified from the traditional sum of squared
residuals.

\texttt{nls.multstart} (Padfield and Matheson (2020)) -- Non-linear least squares regression using
AIC scores with the Levenberg-Marquardt algorithm using multiple starting values
for increasing the chance that the minimum found is the global minimum.

\texttt{nls2} (Grothendieck (2022)) -- Nonlinear least squares by brute force has similar
motivations to \texttt{nls.multstart}, but uses \texttt{nls()} within multiple trials. The author
has extensive expertise in R.

\texttt{nlstac} (Rodriguez-Arias et al. (2020)) -- A set of functions implementing the algorithm described in
Torvisco, Fernandez, and Sanchez (2018) for fitting separable nonlinear regression curves. The
special class of problem for which this package is intended is an important and
difficult one. No vignette is provided, unfortunately.

@@

\texttt{nlraa} (Miguez (2021)) -- a set of nonlinear \emph{selfStart} models, primarily from
agriculture. Most include analytic Jacobian code.

\texttt{optimx} (John C. Nash and Varadhan (2011)) -- This provides
optimizers that can be applied to minimize a nonlinear function which could be
a nonlinear sum of
squares. NOT generally recommended if nonlinear
least squares programs can be easily used, but provides a check and alternative
solvers.

@@

The Hobbs weed infestation problem (John C. Nash (1979, 120)) is a growth curve modeling
task that seems straightforward but is quite nasty. Its very succinct statement
provides the ``short reproducible example'' much requested on R mailing lists.
This is a real problem from a field researcher. The data and graph follow.

\begin{verbatim}
weed <- c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443,
          38.558, 50.156, 62.948, 75.995, 91.972)
tt <- 1:12
weeddf <- data.frame(tt, weed)
plot(weeddf, main="Hobbs weed infestation data")
\end{verbatim}

@@

In reality, there are many problems were \(D\) should not be ignored, but the work
to compute it precisely is considerable. Many work-arounds have been
proposed, of which the Levenberg-Marquardt stabilization
(Levenberg (1944), Marquardt (1963)) is
the most commonly used. For convenience, we will use ``Marquardt'', as we believe he
first incorporated the ideas into a practical computer program.

The usual suggestion is that \(D\) be replaced by a multiple of the unit matrix or else
a multiple of the diagonal part of \(J' J\). In low precision, some elements of
\(J' J\) could underflow to zero (John C. Nash (1977)), and a linear combination of both
choices is an effective compromise. Various choices for
\(D\), as well as a possible line search along the
direction \(\delta\) rather than a unit step (Hartley (1961)),
give rise to several variant
algorithms. ``Marquardt's method'' is a family of methods. Fortunately, most choices
work well.

@@

\begin{verbatim}
# model formulas
frmu <- weed ~ b1/(1+b2*exp(-b3*tt))
frms <- weed ~ 100*c1/(1+10*c2*exp(-0.1*c3*tt))
frmt <- weed ~ Asym /(1 + exp((xmid-tt)/scal))
#
# Starting parameter sets
stu1<-c(b1=1, b2=1, b3=1)
sts1<-c(c1=1, c2=1, c3=1)
stt1<-c(Asym=1, xmid=1, scal=1)
\end{verbatim}

@@

\begin{verbatim}
# Logistic3U
hobbs.res  <-  function(x){ # scaled Hobbs weeds problem -- residual
  # This variant uses looping
  if(length(x) != 3) stop("hobbs.res -- parameter vector n!=3")
  y  <-  c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443, 
           38.558, 50.156, 62.948, 75.995, 91.972)
  tt  <-  1:12
  res  <-  x[1]/(1+x[2]*exp(-x[3]*tt)) - y
}

hobbs.jac  <-  function(x) { # scaled Hobbs weeds problem -- Jacobian
  jj  <-  matrix(0.0, 12, 3)
  tt  <-  1:12
  yy  <-  exp(-x[3]*tt)
  zz  <-  1.0/(1+x[2]*yy)
  jj[tt,1]   <-   zz
  jj[tt,2]   <-   -x[1]*zz*zz*yy
  jj[tt,3]   <-   x[1]*zz*zz*yy*x[2]*tt
  attr(jj, "gradient") <- jj
  jj
}
\end{verbatim}

@@

\begin{verbatim}
#> Error in nls(formula = frmu, start = stu1, data = weeddf) : 
#>   singular gradient
\end{verbatim}

\begin{verbatim}
#> Error in nls(formula = frms, start = sts1, data = weeddf) : 
#>   singular gradient
\end{verbatim}

\begin{verbatim}
#> Error in nls(formula = frmt, start = stt1, data = weeddf) : 
#>   singular gradient
\end{verbatim}
